---
title: "My solution to the challenge"
author: "Gerko Vink"
output: 
   html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: false
---
  
<style type="text/css">
  
body{ /* Normal  */
  font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 18px;
  color: DarkBlue;
}
h1 { /* Header 1 */
  font-size: 18px;
}
h2 { /* Header 2 */
  font-size: 18px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>
  
---

The following packages are required for this practical:
```{r, message=FALSE}
library(dplyr)
library(magrittr)
library(mice)
library(ggplot2)
```

--- 

The basis of the data relations is captured in the variance/covariance matrix and how it translates to data relations. This is most easily linearly summarized in the correlation matrix:
```{r}
mammalsleep %>%
  subset(select = -species) %>% #exclude factor species
  cor(use = "pairwise.complete.obs") #pairwise deletion
```

This matrix contains quite a few cells. To obtain only the correlations with `sws` we could select the respective column:
```{r}
mammalsleep %>%
  subset(select = -species) %>% #exclude factor species
  cor(use = "pairwise.complete.obs") %>% #pairwise deletion
  subset(select = sws) #only column sws from the correlation matrix
```

It seems that the following variables have a rather nice relation with `sws`. As a matter of fact, all predictors do. However, we first focus on this subset:

- `ps`: paradoxical sleep
- `ts` : total sleep
- `gt` : gestation time
- `sei`: sleep exposure index
- `odi`: overall danger index

However, from the larger correlation matrix we can also see that `ts` is highly related to with `ps` - in fact, `ts` is calculated as the sum over `sws` and `ps`. Including both variables will not hurt our $R^2$ per se, but it will certainly trouble the precision of our estimation as including both variables will yield much larger standard errors. I will demonstrate this. For now I choose `ts` as a predictor: `ps` linear association with `sws` is slightly weaker. Moreover, `ps` misses 12 cases, while `ts` only misses 4 cases. 
```{r}
mammalsleep %>%
  summary()
```

Because of the missingness argument it could be highly preferable to use `ts` in the model.

We run the new model:
```{r}
fit <- 
  mammalsleep %$%
  lm(sws ~ ts + gt + sei + odi)
  
fit %>%
  summary()
```
and we have obtained a very high $R^2$, even though only one predictor is significant. If prediction was our goal, we are doing great: the $R^2$ indicates that our predicted outcome correlates $\rho = .969$ with the observed outcome.  

---

## Some further diagnostic checking of this model

---

```{r}
fit %>%
  plot(which = 1:6)
```

Some issues spring to mind:

- There error variance seems to be heteroscedastic [but we have a rather small sample]
- The residuals are not normally distributed in the extreme tails
- The following case has a large leverage: 5
- The following case has large residual: 61
- Case 33 still has a large Cook's distance (no alarming residual or leverage)
```{r}
mammalsleep$species[c(5, 33, 61)]
```

If we sort the data based on `sws`:
```{r}
mammalsleep %>% 
  arrange(desc(sws)) #sort the data in descending order based on sws
```

we see that high `sws` is primarily for lighter mammals without real predatory enemies. 
```{r}
fit %>%
  residuals() 
```

---

## Model with all predictors

---

If we run a model with all predictors, we end up with
```{r}
fit <- 
  mammalsleep %$%
  lm(sws ~ bw + brw + ps + ts + mls + gt + pi + sei + odi) 
  
fit %>%
  summary()
```

This system is linearly dependent; we have the correct answer in the system `sws = ts - ps`. This may not be informative, even though our $R^2$ indicates perfection. In contrast, when we remove `ts`, the $R^2$ naturally drops dramatically. If we remove both `ts` and `ps` a completely different system emerges, but the $R^2$ drops dramatically:
```{r}
fit <- 
  mammalsleep %$%
  lm(sws ~ bw + brw + mls + gt + pi + sei + odi) 
  
fit %>%
  summary()
```

---

## Conclusion

These data are very inter-related. So much so, that it is hard to group all these mammals together in a single linear model. There will always be large residuals: most animals that are really have need quite a lot of food and have no time to eat. Yet, this data set contains only a few of those animals. Other animals are not quite heavy, but still sleep a lot. Take e.g. ourselves (measured in this data set as `man`): We are light, have massive brains and sleep quite a lot - yet we are nowhere near the region in the foodchain any model would expect us to be. 

Bottomline: not all data sets lend themselves for inference based on linear models - at least not without some serious data cleaning and/or feature transformation. That said, it is not hard to find a good fitting and well-performing predictive model in this data set. 

---

End of `Challenge`. 
